Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-12 RevisedSubmission/hines_hetland_2017_gji.tex
did more

Our goal is to find the index set of non-outliers in $\mitbf{d}^*$, which we denote as $\mitbf{\Omega}$. We use a tilde to indicate that an array only contains elements corresponding to $\mitbf{\Omega}$ (e.g., the vector of non-outlier observations is denoted $\tilde{\mitbf{d}}^* = [d_i^*]_{i \in \mitbf{\Omega}}$). The outliers are identified iteratively, and we initiate $\mitbf{\Omega}$ with all $n$ indices. We consider outliers to be data that are poorly explained by the model $\mitbf{G}\mitbf{m} + v(\mitbf{P})$, which is determine by the residual vector 
\begin{align}\label{eq:Residual}
\mitbf{r} &= \mitbf{d}^* - \mathrm{E}\left[\Big(\mitbf{G}\mitbf{m} + v(\mitbf{P})\Big) \Big| \Big(\tilde{\mitbf{d}} = \tilde{\mitbf{d}}^*\Big)\right] \\
          &= \mitbf{d}^*  - 
            \left[\begin{array}{cc}
                  C_v(\mitbf{P},\tilde{\mitbf{P}}) & \mitbf{G} \nonumber \\
                  \end{array}\right]
            \left[\begin{array}{cc}
                  C_v(\tilde{\mitbf{P}},\tilde{\mitbf{P}}) + \mathrm{diag}(\tilde{\mitbf{\sigma}}^2) & \tilde{\mitbf{G}} \\
                  \tilde{\mitbf{G}}^T  & \mitbf{0} \\
                  \end{array}\right]^{-1}
            \left[\begin{array}{c}
                  \tilde{\mitbf{d}}^* \\
                  \mitbf{0} \\
                  \end{array}\right].
\end{align}
Data with abnormally large residuals are identified as outliers. For each iteration, we compute $\mitbf{r}$ and then update $\mitbf{\Omega}$ so that it contains the indices of $\mitbf{r}$ whose weighted values are less than $\lambda$ times the weighted root mean square of $\tilde{\mitbf{r}}$, 
\begin{equation}\label{eq:Update}
\mitbf{\Omega} \leftarrow \left\{i : \left|\frac{r_i}{\sigma_i}\right| < \lambda \cdot \sqrt{\frac{1}{|\mitbf{\Omega}|} \sum_{j \in \mitbf{\Omega}} \frac{r_j^2}{\sigma_j^2}} \right\}.
\end{equation} 
Iterations continue until the new $\mitbf{\Omega}$ is the same as the previous $\mitbf{\Omega}$. We use $\lambda=4$ in this study, which in our experience accurately identifies outliers without unnecessarily decimating the data.

The outlier detection algorithm is demonstrated in Figure \ref{fig:Outliers}. In the demonstration we use the easting component of displacements at a single station, SC03, which is located on Mt. Olympus in Washington state. Station SC03 records anomalous observations during the winter, presumably because of snow and ice accumulation, and we want to remove these observations. The station also records periodic westward motion from slow slip events, and we want to keep this deformation intact. The detected outliers are shown in (Panel B). For comparison, we also show the detected outliers when we do not include the Gaussian process $v$ in our model for the data (Panel A). When $v$ is not included, real transient deformation resulting from slow slip events is erroneously identified as outliers. When $v$ is included, the identified outliers only consist of the anomalous deformation that lacks temporal continuity.

>>>>

Our goal is to find the index set of non-outliers in $\mitbf{d}^*$, which we denote as $\mitbf{\Omega}$. We use a tilde to indicate that an array only contains elements corresponding to $\mitbf{\Omega}$ (e.g., the vector of non-outlier observations is denoted $\tilde{\mitbf{d}}^* = [d_i^*]_{i \in \mitbf{\Omega}}$). The outliers are identified iteratively, and we initiate $\mitbf{\Omega}$ with all $n$ indices. We consider outliers to be data that are poorly explained by the model $\mitbf{G}\mitbf{m} + v(\mitbf{P})$, which is determine by the residual vector 
\begin{align}\label{eq:Residual}
\mitbf{r} &= \mitbf{d}^* - \mathrm{E}\left[\Big(\mitbf{G}\mitbf{m} + v(\mitbf{P})\Big) \Big| \Big(\tilde{\mitbf{d}} = \tilde{\mitbf{d}}^*\Big)\right] \\
          &= \mitbf{d}^*  - 
            \left[\begin{array}{cc}
                  C_v(\mitbf{P},\tilde{\mitbf{P}}) & \mitbf{G} \nonumber \\
                  \end{array}\right]
            \left[\begin{array}{cc}
                  C_v(\tilde{\mitbf{P}},\tilde{\mitbf{P}}) + \mathrm{diag}(\tilde{\mitbf{\sigma}}^2) & \tilde{\mitbf{G}} \\
                  \tilde{\mitbf{G}}^T  & \mitbf{0} \\
                  \end{array}\right]^{-1}
            \left[\begin{array}{c}
                  \tilde{\mitbf{d}}^* \\
                  \mitbf{0} \\
                  \end{array}\right].
\end{align}
Data with abnormally large residuals are identified as outliers. For each iteration, we compute $\mitbf{r}$ and then update $\mitbf{\Omega}$ so that it contains the indices of $\mitbf{r}$ whose weighted values are less than $\lambda$ times the weighted root mean square of $\tilde{\mitbf{r}}$, 
\begin{equation}\label{eq:Update}
\mitbf{\Omega} \leftarrow \left\{i : \left|\frac{r_i}{\sigma_i}\right| < \lambda \cdot \sqrt{\frac{1}{|\mitbf{\Omega}|} \sum_{j \in \mitbf{\Omega}} \frac{r_j^2}{\sigma_j^2}} \right\}.
\end{equation} 
Iterations continue until the new $\mitbf{\Omega}$ is the same as the previous $\mitbf{\Omega}$.

The outlier detection algorithm is demonstrated in Figure \ref{fig:Outliers}. For the demonstration, we use the easting component of displacements at a single station, SC03, which is located on Mt. Olympus in Washington state. Station SC03 records anomalous observations during the winter, presumably because of snow and ice accumulation, and we want to remove these observations. The station also records periodic westward motion from slow slip events, and we want to keep this deformation intact. The detected outliers are shown in Panel B. For comparison, we also show the detected outliers when we do not include the Gaussian process $v$ in our model for the data (Panel A). When $v$ is not included, real transient deformation resulting from slow slip events is erroneously identified as outliers. When $v$ is included, the identified outliers only consist of the anomalous deformation that lacks temporal continuity.  It should be noted that we use $\lambda = 2.5$ for this demonstration, which causes the outlier detection algorithm to be particularly aggressive. In Section \ref{sec:Cascadia}, we clean the data using a more tolerant $\lambda = 4.0$.
